---
title: "The (Residual) Monopolist's Best Response"
author: "Reimplementation of Flavio Toxvaerd's Model"
format: 
  html:
    page-layout: full
    theme: cosmo
server: shiny
---

## Model Overview

This interactive graph illustrates the **Residual Monopolist's Problem**. It shows the profits of a Cournot firm ($i$) given the production level of rival firms ($j$).

**The Economic Intuition:**
The problem of a Cournot firm is similar in nature to that of a monopolist, except that the demand function the firm faces is the **residual demand function** rather than the underlying total market demand[cite: 79].

### Experiments
1.  [cite_start]**Shift the Profit Curve:** By increasing the production of other firms ($q_j$), observe how the profit function of the firm shifts leftward in the upper panel[cite: 80].
2.  **Analyze Optimal Output:** Notice how that shift changes the optimal production level ($q_i$) of the firm.
3.  [cite_start]**Explain:** Use the figure to explain why an increase in the rival's output causes the firm to decrease its output[cite: 84].

```{r}
#| context: setup
library(shiny)
library(ggplot2)
library(latex2exp)

# Define plotting theme for consistency
theme_econ <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "italic"),
    panel.grid.minor = element_blank()
  )

```


```{r}
#| panel: sidebar

# Controls based on the "Controls" table in the PDF
sliderInput(
  "a",
  label = "Demand Intercept (a):",
  min = 15,
  max = 20,
  value = 15,
  step = 0.5
)

sliderInput(
  "b",
  label = "Demand Slope (b):",
  min = 0.75,
  max = 5,
  value = 0.75,
  step = 0.25
)

sliderInput(
  "ci",
  label = "Marginal Cost (c):",
  min = 2,
  max = 5,
  value = 2,
  step = 0.5
)

sliderInput(
  "qj",
  label = "Rival Output (qj):",
  min = 0,
  max = 20,
  value = 6.4,
  step = 0.1
)

hr()
helpText("Original concept by Flavio Toxvaerd (University of Cambridge).")
```


```{r}
#| panel: fill

# Layout with Description Tab and Plot Tab for cleaner UI
tabsetPanel(
  tabPanel(
    "Interactive Graphs",
    br(),
    plotOutput("profitPlot", height = "350px"),
    plotOutput("brPlot", height = "350px")
  ),
  tabPanel(
    "Mathematical Details",
    withMathJax(),
    h3("The Model"),
    p("The market is characterized by the inverse demand function:"),
    p("$$p = a - b(q_i + q_j)$$"),
    p("Where:"),
    tags$ul(
      tags$li("\\(q_i\\) is the quantity of the firm."),
      tags$li("\\(q_j\\) is the quantity produced by the firm's rivals."),
      tags$li("\\(a\\) and \\(b\\) are demand parameters.")
    ),
    p("The firm has constant marginal cost \\(c\\)."),
    h4("The Panels"),
    p(
      strong("Upper Panel:"),
      "Shows the profit function. Increasing \\(q_j\\) shifts this curve left."
    ),
    p(
      strong("Lower Panel:"),
      "Plots the different combinations of \\(q_j\\) and the accompanying (privately) optimal production \\(q_i\\). This curve traces out the Cournot firm's Best Response (or reaction function)."
    )
  )
)
```


```{r}
#| context: server

# Reactive Optimization Calculation
# Calculates optimal qi given the current inputs
opt_vals <- reactive({
  a <- input$a
  b <- input$b
  ci <- input$ci
  qj <- input$qj

  # Optimal qi (Reaction to qj)
  # Derived from FOC of Profit = qi * (a - ci - b(qi + qj))
  qi_opt <- (a - ci - b * qj) / (2 * b)

  # Max Profit at that qi
  pi_max <- qi_opt * (a - ci - b * (qi_opt + qj))

  # Monopoly Reference (if qj were 0)
  qi_mon <- (a - ci) / (2 * b)
  pi_mon <- qi_mon * (a - ci - b * qi_mon)

  # Clamp negative quantities to 0 for display safety
  if (qi_opt < 0) {
    qi_opt <- 0
    pi_max <- 0
  }

  list(qi = qi_opt, pi = pi_max, qi_mon = qi_mon, pi_mon = pi_mon)
})

# Plot 1: The Profit Function (Parabola)
output$profitPlot <- renderPlot({
  a <- input$a
  b <- input$b
  ci <- input$ci
  qj <- input$qj
  vals <- opt_vals()

  # Create data for curves
  qi_seq <- seq(0, 20, length.out = 300)

  # Current Profit Curve (Blue)
  pi_curr <- qi_seq * (a - ci - b * (qi_seq + qj))

  # Monopoly Reference Curve (Dashed Black)
  pi_ref <- qi_seq * (a - ci - b * qi_seq)

  df <- data.frame(qi = qi_seq, pi_curr = pi_curr, pi_ref = pi_ref)

  ggplot(df, aes(x = qi)) +
    # Reference Curve (qj = 0)
    geom_line(
      aes(y = pi_ref),
      linetype = "dashed",
      color = "black",
      alpha = 0.6
    ) +
    # Current Curve (Actual qj)
    geom_line(aes(y = pi_curr), color = "blue", size = 1.2) +

    # Points and Drop lines for Optimal
    annotate("point", x = vals$qi, y = vals$pi, color = "blue", size = 4) +
    annotate(
      "segment",
      x = vals$qi,
      xend = vals$qi,
      y = 0,
      yend = vals$pi,
      linetype = "dotted",
      color = "blue"
    ) +

    # Points and Drop lines for Monopoly Reference
    annotate(
      "point",
      x = vals$qi_mon,
      y = vals$pi_mon,
      color = "black",
      size = 2,
      alpha = 0.5
    ) +
    annotate(
      "segment",
      x = vals$qi_mon,
      xend = vals$qi_mon,
      y = 0,
      yend = vals$pi_mon,
      linetype = "dotted",
      color = "black",
      alpha = 0.5
    ) +

    # Labels and Limits
    labs(
      title = "Residual Monopolist's Profit",
      x = TeX("$q_i$"),
      y = TeX("$\\pi_i(q_i, q_j)$")
    ) +
    coord_cartesian(ylim = c(-10, 75), xlim = c(0, 20)) +
    theme_econ
})

# Plot 2: Best Response Function
output$brPlot <- renderPlot({
  a <- input$a
  b <- input$b
  ci <- input$ci
  qj_curr <- input$qj
  qi_opt <- opt_vals()$qi

  # The Best Response Function: qi = (a - ci - b*qj) / 2b
  # Rearrange for plotting qj on Y: qj = (a - ci)/b - 2*qi

  qi_seq <- seq(0, 20, length.out = 100)
  qj_line <- (a - ci) / b - 2 * qi_seq

  df_br <- data.frame(qi = qi_seq, qj = qj_line)
  df_br <- df_br[df_br$qj >= 0, ]

  ggplot() +
    # The Best Response Line
    geom_line(data = df_br, aes(x = qi, y = qj), size = 1.2) +

    # Current State Point
    annotate("point", x = qi_opt, y = qj_curr, color = "blue", size = 4) +

    # Drop lines to axes
    annotate(
      "segment",
      x = qi_opt,
      xend = qi_opt,
      y = 0,
      yend = qj_curr,
      linetype = "dashed",
      color = "gray"
    ) +
    annotate(
      "segment",
      x = 0,
      xend = qi_opt,
      y = qj_curr,
      yend = qj_curr,
      linetype = "dashed",
      color = "gray"
    ) +

    labs(
      title = "Best Response Function",
      x = TeX("$q_i$"),
      y = TeX("$q_j$ (Rival Output)")
    ) +
    coord_cartesian(ylim = c(0, 20), xlim = c(0, 20)) +
    theme_econ
})
```

