---
title: "IO Diagrams (TikZ embedded)"
format:
  revealjs:
    slide-number: true
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r}
#| label: tikz-helper
#| include: false

library(here)
library(knitr)

dir.create(
  here("lecture-slides", "figs"),
  showWarnings = FALSE,
  recursive = TRUE
)

# Render TikZ -> PDF via pdflatex, then PDF -> SVG via Inkscape,
# optionally clean auxiliary files, then return an <img> tag
render_tikz_svg <- function(
  tikz_picture,
  name,
  width = "70%",
  engine = "pdflatex",
  cleanup = TRUE,
  keep_pdf = FALSE
) {
  figs_dir <- here("lecture-slides", "figs")

  tex_file <- file.path(figs_dir, paste0(name, ".tex"))
  pdf_file <- file.path(figs_dir, paste0(name, ".pdf"))
  svg_file <- file.path(figs_dir, paste0(name, ".svg"))

  # Standalone doc for tight cropping
  tex <- paste0(
    "\\documentclass[tikz,border=2pt]{standalone}\n",
    "\\usepackage{pgfplots}\n",
    "\\pgfplotsset{compat=1.18}\n",
    "\\begin{document}\n",
    tikz_picture,
    "\n\\end{document}\n"
  )
  writeLines(tex, tex_file)

  # Compile to PDF
  if (Sys.which(engine) == "") {
    stop("LaTeX engine not found on PATH: ", engine)
  }

  res <- system2(
    engine,
    c(
      "-interaction=nonstopmode",
      paste0("--output-directory=", shQuote(figs_dir)),
      shQuote(tex_file)
    ),
    stdout = TRUE,
    stderr = TRUE
  )

  if (!file.exists(pdf_file)) {
    stop("PDF not produced. LaTeX output:\n", paste(res, collapse = "\n"))
  }

  # Convert PDF -> SVG using Inkscape
  inkscape <- Sys.which("inkscape")
  if (inkscape == "") {
    stop("Inkscape not found on PATH.")
  }

  system2(
    inkscape,
    c(
      shQuote(pdf_file),
      "--export-type=svg",
      paste0("--export-filename=", shQuote(svg_file))
    )
  )

  if (!file.exists(svg_file)) {
    stop("SVG not produced by Inkscape.")
  }

  # ---- Cleanup (only after success) ----
  if (cleanup) {
    aux_ext <- c(
      ".aux",
      ".log",
      ".out",
      ".toc",
      ".synctex.gz",
      ".fls",
      ".fdb_latexmk",
      ".tex"
    )

    aux_files <- file.path(figs_dir, paste0(name, aux_ext))
    aux_files <- aux_files[file.exists(aux_files)]

    if (length(aux_files) > 0) {
      unlink(aux_files)
    }

    if (!keep_pdf && file.exists(pdf_file)) {
      unlink(pdf_file)
    }
  }

  # Embed as HTML
  knitr::asis_output(sprintf(
    '<img src="%s" style="width:%s; height:auto;">',
    svg_file,
    width
  ))
}

```


```{r}
#| label: fig-hotelling-fixed-location
render_tikz_svg(
  tikz_picture = "
\\begin{tikzpicture}[>=stealth, line cap=round]

% --- Key x-positions (edit these to change spacing) ---
\\def\\xO{0}
\\def\\xFone{3}
\\def\\xMC{6}
\\def\\xFtwo{9}
\\def\\xL{12}

% --- Baseline ---
\\draw[thick] (\\xO,0) -- (\\xL,0);

% --- Points ---
\\fill (\\xO,0) circle (2pt);
\\fill (\\xFone,0) circle (2pt);
\\fill (\\xMC,0) circle (2pt);
\\fill (\\xFtwo,0) circle (2pt);
\\fill (\\xL,0) circle (2pt);

% --- End labels ---
\\node[left]  at (\\xO,0) {$0$};
\\node[right] at (\\xL,0) {$L$};

% --- Braces above segments: a, x, y, b ---
\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xO,0.55) -- (\\xFone,0.55) node[midway, yshift=10pt] {$a$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xFone,0.55) -- (\\xMC,0.55) node[midway, yshift=10pt] {$x$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xMC,0.55) -- (\\xFtwo,0.55) node[midway, yshift=10pt] {$y$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xFtwo,0.55) -- (\\xL,0.55) node[midway, yshift=10pt] {$b$};

% --- Firm labels with downward arrows ---
\\node at (\\xFone,1.45) {Firm 1};
\\draw[->] (\\xFone,1.20) -- (\\xFone,0.12);

\\node at (\\xFtwo,1.45) {Firm 2};
\\draw[->] (\\xFtwo,1.20) -- (\\xFtwo,0.12);

% --- Marginal consumer label with upward arrow from text to point ---
\\node[align=center] at (\\xMC,-1.15) {Marginal\\\\consumer};
\\draw[->] (\\xMC,-0.75) -- (\\xMC,-0.08);

\\end{tikzpicture}",
  name = "fig-hotelling-fixed-location",
  width = "70%"
)
```

```{r}
#| label: fig-hotelling-movement-location
render_tikz_svg(
  tikz_picture = "
\\begin{tikzpicture}[>=stealth, line cap=round]

% --- Key x-positions (edit to change spacing) ---
\\def\\xO{0}
\\def\\xM{6}   % marginal consumer / co-location point of firms 1 and 2
\\def\\xL{12}

% --- Baseline ---
\\draw[thick] (\\xO,0) -- (\\xL,0);

% --- Endpoints and midpoint ---
\\fill (\\xO,0) circle (2pt);
\\fill (\\xM,0) circle (2pt);
\\fill (\\xL,0) circle (2pt);

% --- End labels ---
\\node[left]  at (\\xO,0) {$0$};
\\node[right] at (\\xL,0) {$L$};

% --- Braces above segments: a and b ---
\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xO,0.55) -- (\\xM,0.55) node[midway, yshift=10pt] {$a$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xM,0.55) -- (\\xL,0.55) node[midway, yshift=10pt] {$b$};

% --- Firms label with downward arrow to midpoint ---
\\node at (\\xM,1.45) {Firms 1 and 2};
\\draw[->] (\\xM,1.20) -- (\\xM,0.12);

% --- Marginal consumer label with upward arrow to midpoint ---
\\node[align=center] at (\\xM,-1.15) {Marginal consumer};
\\draw[->] (\\xM,-0.75) -- (\\xM,-0.08);

\\end{tikzpicture}
",
  name = "fig-hotelling-movement-location",
  width = "70%"
)
```


```{r}
#| label: fig-hotelling-profit-function

render_tikz_svg(
  tikz_picture = "
\\begin{tikzpicture}
\\begin{axis}[
  width=12.8cm, height=6.0cm,
  xmin=0, xmax=10.6,
  ymin=0, ymax=5.2,
  axis lines=left,
  axis line style={thick, -{Stealth[length=2.2mm]}},
  ticks=none,
  clip=false,
  xlabel={$p_1$},
  ylabel={$\\pi_1(p_1,\\bar p_2)$},
  xlabel style={at={(axis description cs:1.02,0.02)}, anchor=west},
  ylabel style={at={(axis description cs:0.02,1.02)}, anchor=south},
]

% --- Key x locations (chosen to match the screenshot spacing) ---
\\pgfmathsetmacro{\\xL}{3.25}  % left cutoff: \\bar p_2 - c(L-a-b)
\\pgfmathsetmacro{\\xM}{6.20}  % middle cutoff: \\bar p_1
\\pgfmathsetmacro{\\xR}{8.35}  % right cutoff: \\bar p_2 + c(L-a-b)

% --- Key y levels (enforce: yTop > yPeak) ---
\\pgfmathsetmacro{\\yTop}{4.70}  % highest point (end of linear segment)
\\pgfmathsetmacro{\\yDrop}{2.55} % value after discrete drop
\\pgfmathsetmacro{\\yPeak}{3.85} % peak of middle segment (STRICTLY below yTop)

% --- Curvature parameter k from vertex form so that parabola passes through (xL,yDrop) ---
% y(x) = yPeak - k (x-xM)^2, with y(xL)=yDrop => k=(yPeak-yDrop)/(xL-xM)^2
\\pgfmathsetmacro{\\k}{(\\yPeak-\\yDrop)/((\\xL-\\xM)^2)}

% --- Implied right-end y-value (consistent with the same parabola) ---
\\pgfmathsetmacro{\\yR}{\\yPeak - \\k*(\\xR-\\xM)^2}

% 1) Left linear segment (formula): line from (0,0) to (xL,yTop)
\\addplot[thick, domain=0:\\xL, samples=2] ({x},{(\\yTop/\\xL)*x});

% 2) Discrete drop at xL
\\addplot[thick] coordinates {(\\xL,\\yTop) (\\xL,\\yDrop)};

% 3) Middle segment (formula): concave parabola in vertex form
\\addplot[thick, domain=\\xL:\\xR, samples=200] ({x},{\\yPeak - \\k*(x-\\xM)^2});

% --- Dashed vertical cutoffs (match slide: extend to each segment’s value) ---
\\addplot[densely dashed, thick, gray] coordinates {(\\xL,0) (\\xL,\\yTop)};
\\addplot[densely dashed, thick, gray] coordinates {(\\xM,0) (\\xM,\\yPeak)};
\\addplot[densely dashed, thick, gray] coordinates {(\\xR,0) (\\xR,\\yR)};

% --- x-axis labels ---
\\node[anchor=north] at (axis cs:\\xL,0) {$\\bar p_2-c(L-a-b)$};
\\node[anchor=north] at (axis cs:\\xM,0) {$\\bar p_1$};
\\node[anchor=north] at (axis cs:\\xR,0) {$\\bar p_2+c(L-a-b)$};

\\end{axis}
\\end{tikzpicture}
",
  name = "fig-hotelling-profit-function",
  width = "70%"
)

```


```{r}
#| label: fig-salop-cycle

render_tikz_svg(
  tikz_picture = "
\\begin{tikzpicture}[x=1cm,y=1cm, line cap=round, line join=round]

% --- Parameters ---
\\def\\R{2.2} % radius

% --- Circle ---
\\draw[thick] (0,0) circle (\\R);

% --- Four points on the circle (N, E, S, W) ---
\\fill (0,\\R) circle (1.6pt);
\\fill (\\R,0) circle (1.6pt);
\\fill (0,-\\R) circle (1.6pt);
\\fill (-\\R,0) circle (1.6pt);

% --- Labels at points ---
\\node[above] at (0,\\R) {$\\bar p$};
\\node[right] at (\\R,0) {$\\bar p$};
\\node[below] at (0,-\\R) {$p$};
\\node[left]  at (-\\R,0) {$\\bar p$};

% --- Quarter-arc labels (placed in each quadrant) ---
\\node at (-1.75,  1.75) {$\\frac{1}{4}$};
\\node at ( 1.75,  1.75) {$\\frac{1}{4}$};
\\node at (-1.75, -1.75) {$\\frac{1}{4}$};
\\node at ( 1.75, -1.75) {$\\frac{1}{4}$};

\\end{tikzpicture}
",
  name = "fig-salop-cycle",
  width = "70%"
)

```



```{r}
#| label: fig-salop-line
render_tikz_svg(
  tikz_picture = "
\\begin{tikzpicture}[>=stealth, line cap=round]

% --- Key positions ---
\\def\\xL{-4.5}
\\def\\xC{0}
\\def\\xR{4.5}

% --- Baseline with arrows ---
\\draw[thick, <->] (-5.5,0) -- (5.5,0);

% --- Points ---
\\fill (\\xL,0) circle (2pt);
\\fill (\\xC,0) circle (2pt);
\\fill (\\xR,0) circle (2pt);

% --- Point labels ---
\\node[below] at (\\xL,0) {$\\bar p$};
\\node[below] at (\\xC,0) {$p$};
\\node[below] at (\\xR,0) {$\\bar p$};

% --- Braces for distances ---
\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xL,0.6) -- (\\xC,0.6)
  node[midway, yshift=10pt] {$\\frac{1}{n}$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xC,0.6) -- (\\xR,0.6)
  node[midway, yshift=10pt] {$\\frac{1}{n}$};

\\end{tikzpicture}",
  name = "fig-salop-line",
  width = "70%"
)

```



```{r}
#| label: salop-line-mod

render_tikz_svg(
  tikz_picture = "
\\begin{tikzpicture}[>=stealth, line cap=round]

% --- Key positions ---
\\def\\xL{-5.0}   % left \\bar p
\\def\\xA{-2.0}   % left marginal consumer
\\def\\xC{0.0}    % p (center)
\\def\\xB{2.0}    % right marginal consumer
\\def\\xR{5.0}    % right \\bar p

% --- Baseline with arrows ---
\\draw[thick, <->] (-6.0,0) -- (6.0,0);

% --- Points ---
\\fill (\\xL,0) circle (2pt);
\\fill (\\xA,0) circle (2pt);
\\fill (\\xC,0) circle (2pt);
\\fill (\\xB,0) circle (2pt);
\\fill (\\xR,0) circle (2pt);

% --- Point labels ---
\\node[below] at (\\xL,0) {$\\bar p$};
\\node[below] at (\\xC,0) {$p$};
\\node[below] at (\\xR,0) {$\\bar p$};

% --- Braces for d_m on each side of p ---
\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xA,0.65) -- (\\xC,0.65)
  node[midway, yshift=10pt] {$d_m$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xC,0.65) -- (\\xB,0.65)
  node[midway, yshift=10pt] {$d_m$};

% --- Marginal consumer labels with downward arrows to points xA and xB ---
\\node[align=center] at (\\xA,1.55) {Marginal\\\\consumer};
\\draw[->, thick] (\\xA,1.20) -- (\\xA,0.12);

\\node[align=center] at (\\xB,1.55) {Marginal\\\\consumer};
\\draw[->, thick] (\\xB,1.20) -- (\\xB,0.12);

% --- Net surplus < 0 annotations with arrows pointing outward ---
\\node[align=center] at (-3.9,0.95) {Net surplus $< 0$};
\\draw[->, thick] (-3.2,0.55) -- (-4.9,0.55);

\\node[align=center] at (3.9,0.95) {Net surplus $< 0$};
\\draw[->, thick] (3.2,0.55) -- (4.9,0.55);

\\end{tikzpicture}",
  name = "fig-salop-line-mod",
  width = "70%"
)

```


```{r}
#| label: fig-salop-line-comp-region

render_tikz_svg(
  tikz_picture = "
\\begin{tikzpicture}[>=stealth, line cap=round]

% --- Key positions ---
\\def\\xL{-5.0}   % left \\bar p
\\def\\xA{-2.0}   % left marginal consumer
\\def\\xC{0.0}    % p (center)
\\def\\xB{2.0}    % right marginal consumer
\\def\\xR{5.0}    % right \\bar p

% --- Baseline with arrows ---
\\draw[thick, <->] (-6.0,0) -- (6.0,0);

% --- Points ---
\\fill (\\xL,0) circle (2pt);
\\fill (\\xA,0) circle (2pt);
\\fill (\\xC,0) circle (2pt);
\\fill (\\xB,0) circle (2pt);
\\fill (\\xR,0) circle (2pt);

% --- Point labels ---
\\node[below] at (\\xL,0) {$\\bar p$};
\\node[below] at (\\xC,0) {$p$};
\\node[below] at (\\xR,0) {$\\bar p$};

% --- Braces: (1/n - x) on the outer segments, x on the inner segments ---
\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xL,0.65) -- (\\xA,0.65)
  node[midway, yshift=10pt] {$\\frac{1}{n}-x$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xA,0.65) -- (\\xC,0.65)
  node[midway, yshift=10pt] {$x$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xC,0.65) -- (\\xB,0.65)
  node[midway, yshift=10pt] {$x$};

\\draw[decorate, decoration={brace, amplitude=4pt}]
  (\\xB,0.65) -- (\\xR,0.65)
  node[midway, yshift=10pt] {$\\frac{1}{n}-x$};

% --- Marginal consumer labels with downward arrows to points xA and xB ---
\\node[align=center] at (\\xA,1.55) {Marginal\\\\consumer};
\\draw[->, thick] (\\xA,1.20) -- (\\xA,0.12);

\\node[align=center] at (\\xB,1.55) {Marginal\\\\consumer};
\\draw[->, thick] (\\xB,1.20) -- (\\xB,0.12);

\\end{tikzpicture}",
  name = "fig-salop-line-comp-region",
  width = "70%"
)

```


```{r}
#| label: fig-salop-demand-comp-mon
render_tikz_svg(
  tikz_picture = "
\\begin{tikzpicture}[>=stealth, line cap=round, line join=round]

% --- Axes ---
\\draw[->, thick] (0,0) -- (6.6,0) node[anchor=west] {$q$};
\\draw[->, thick] (0,0) -- (0,5.2) node[anchor=south] {$p$};

% --- Mark on p-axis ---
\\fill (0,4.2) circle (1.5pt);
\\node[left] at (0,4.2) {$v$};

% --- Demand / kinked line ---
\\draw[thick] (0,4.2) -- (3.0,3.0) -- (5.2,1.6);

% --- Vertical dashed separator ---
\\draw[densely dashed, thick] (3.0,0) -- (3.0,4.8);

% --- Region labels ---
\\node at (1.9,4.6) {Monopoly};
\\node at (1.9,4.2) {region};

\\node at (4.2,4.6) {Competitive};
\\node at (4.2,4.2) {region};

\\end{tikzpicture}",
  name = "fig-salop-demand-comp-mon",
  width = "70%"
)

```

```{r}
#| label: fig-stackelberg-br

render_tikz_svg(
  tikz_picture = "\\begin{tikzpicture}
\\begin{axis}[
  width=13cm,height=8cm,
  xmin=0,xmax=10,
  ymin=0,ymax=10,
  axis lines=left,
  ticks=none,
  xlabel={$q_1$},
  ylabel={$q_2$},
  xlabel style={at={(axis description cs:0.98,0.02)}, anchor=west},
  ylabel style={at={(axis description cs:0.02,0.98)}, anchor=south},
  clip=false,
]

% --- Choose parameters just for geometry (keeps everything in-frame) ---
% We target BR2(q1) = alpha - beta*q1 with intercept 8 and slope -0.5
\\pgfmathsetmacro{\\alpha}{8}
\\pgfmathsetmacro{\\beta}{0.5}

% Best response of firm 2
\\addplot[thick, domain=0:10, samples=2] ({x},{max(0,\\alpha - \\beta*x)});
\\node[anchor=west] at (axis cs:7.2,4.4) {$BR_2(q_1)$};

% Cournot point (on BR with symmetry in the linear case)
% Solve q = alpha/(1+beta)??? Here we set it explicitly to match slide geometry
\\pgfmathsetmacro{\\qC}{\\alpha/(1+2*\\beta)} % (consistent with linear model mapping)
\\pgfmathsetmacro{\\q2C}{\\alpha - \\beta*\\qC}

% Stackelberg point: q1 larger, q2 on BR
\\pgfmathsetmacro{\\qS}{2*\\qC}             % leader expands output
\\pgfmathsetmacro{\\q2S}{\\alpha - \\beta*\\qS}

% Mark points
\\addplot[only marks, mark=*, mark size=1.8pt] coordinates {(\\qC,\\q2C)};
\\node[anchor=south west] at (axis cs:\\qC,\\q2C) {Cournot};

\\addplot[only marks, mark=*, mark size=1.8pt] coordinates {(\\qS,\\q2S)};
\\node[anchor=south west] at (axis cs:\\qS,\\q2S) {Stackelberg};

% Dotted guides
\\addplot[densely dotted] coordinates {(\\qC,0) (\\qC,\\q2C)};
\\addplot[densely dotted] coordinates {(0,\\q2C) (\\qC,\\q2C)};
\\addplot[densely dotted] coordinates {(\\qS,0) (\\qS,\\q2S)};
\\addplot[densely dotted] coordinates {(0,\\q2S) (\\qS,\\q2S)};

% --- Leader isoprofit curves (stylized; computed from linear inverse demand) ---
% Take linear demand P = A - (q1+q2), MC=0 purely for shape.
% Leader profit: pi1 = q1*(A - q1 - q2)
% => isoprofit: q2 = A - q1 - pi/q1
\\pgfmathsetmacro{\\A}{10}
\\addplot[thin, domain=0.7:9.5, samples=200] ({x},{max(0,\\A - x - 10/x)});
\\addplot[thin, domain=0.7:9.5, samples=200] ({x},{max(0,\\A - x - 14/x)});
\\addplot[thin, domain=0.7:9.5, samples=200] ({x},{max(0,\\A - x - 18/x)});

\\node[anchor=west] at (axis cs:6.9,1.2) {Leader isoprofits};

% Optional label to stress strategic substitutes (downward BR)
\\node[anchor=west] at (axis cs:0.4,9.4) {Strategic substitutes: $BR_2'(q_1)<0$};

\\end{axis}
\\end{tikzpicture}
",
  name = "fig-stackelberg-br",
  width = "70%"
)