---
title: "EC5223 – Industrial Economics"
subtitle: "Lecture 3: Economics of Innovation"
format:
  revealjs:
    slide-number: true
    css: lecture-slides/theme/st-andrews.scss
    theme: default
    chalkboard: true
    preview-links: auto
    transition: none
    background-transition: fade
    width: 1600
    height: 900
    margin: 0.1
---

```{r}
#| label: setup
library(ggplot2)
library(dplyr)
```

## Innovation: overview

Why study innovation?

* Innovation is a **stochastic** process.
* The **probability of success** depends on resources allocated.

Types:

* **Process innovation**: shifts the cost function (typically lowers marginal cost).
* **Product innovation**: shifts the demand function.

Stages:

* Research
* Development
* Adoption

## Innovation and competition: two-way relation

Two-way relationship:

* How **market structure** affects incentives to innovate

* How **innovation** affects market structure

## Market structure affects incentives to innovate

Compare maximum willingness to pay (WTP) for an innovation across:

1. Monopoly (before and after)
2. Perfect competition (before; monopoly rights after)
3. Social planner

Interpretation:

* Max WTP = **increase in profit** (firm) or **increase in welfare** (planner) from acquiring the innovation.

## Example setup

Market demand (inverse):

$$
P(Q) = A - Q
$$

Process innovation:

* Marginal cost falls from $c_0$ to $c_1$
* Parameters satisfy $c_1 < c_0 < A$

## Example: monopoly (before vs after)

Monopoly chooses $Q$ to maximize
$$
\pi(Q) = (A - Q - c)Q.
$$

FOC:
$$
A - 2Q - c = 0 \quad \Rightarrow \quad Q^m(c) = \frac{A-c}{2}.
$$

Price:
$$
P^m(c) = A - Q^m(c) = \frac{A+c}{2}.
$$

Profit:
$$
\pi^m(c) = (P^m(c) - c)Q^m(c) = \frac{(A-c)^2}{4}.
$$

So the monopolist’s max WTP for the innovation is:
$$
\Delta \pi^m = \pi^m(c_1) - \pi^m(c_0).
$$

## Example: perfect competition (before; monopoly rights after)

Before innovation (all firms at MC $c_0$):

* Competitive price: $P^{pc}_0 = c_0$
* Representative firm earns (approximately) zero economic profit.

After innovation (innovator has exclusive use; rivals remain at $c_0$):

* If innovation is **drastic**, innovator can price like a monopolist at cost $c_1$ without being constrained by rivals.
* If **non-drastic**, innovator is constrained by the competitive fringe at $c_0$.

Key comparison (from the lecture logic):

* A competitive firm can be willing to pay **more** than a monopolist because it is “creating” profit rather than replacing an existing one.

## Size of (process) innovation

Definitions:

* **Drastic (major) innovation**: innovator can behave as an unconstrained monopolist.
* **Non-drastic (minor) innovation**: innovator has a cost advantage but remains constrained by rivals.

Criterion (as stated in the lecture):

* Drastic if the monopoly price at cost $c_1$ falls below rivals’ cost $c_0$:
  $$
  P^m(c_1) < c_0.
  $$

Using $P^m(c) = (A+c)/2$:
$$
\frac{A+c_1}{2} < c_0.
$$

## Figure: Drastic vs non-drastic process innovation (recreated)

<!-- ```{r}
#| label: fig-drastic-nondrastic
#| fig-cap: "Size of (process) innovation: drastic vs non-drastic (recreation of lecture diagram)."
# Parameters (choose values consistent with the lecture inequalities)
A <- 10

# Left panel: drastic example
c0_L <- 6
c1_L <- 1

# Right panel: non-drastic example
c0_R <- 6
c1_R <- 4.5

make_panel <- function(c0, c1, title) {
  Qmax <- A
  df <- tibble(Q = seq(0, Qmax, length.out = 400)) |>
    mutate(
      P = A - Q,
      MR = A - 2 * Q
    )

  # Monopoly with cost c1
  Qm1 <- (A - c1) / 2
  Pm1 <- (A + c1) / 2

  ggplot(df, aes(x = Q)) +
    geom_line(aes(y = P), linewidth = 1) +
    geom_line(aes(y = MR), linewidth = 0.9, linetype = "dashed") +
    geom_hline(yintercept = c0, linewidth = 0.8) +
    geom_hline(yintercept = c1, linewidth = 0.8) +
    geom_vline(xintercept = Qm1, linetype = "dotted") +
    geom_hline(yintercept = Pm1, linetype = "dotted") +
    annotate("text", x = Qmax * 0.72, y = A - Qmax * 0.72 + 0.4, label = "D") +
    annotate(
      "text",
      x = Qmax * 0.72,
      y = A - 2 * (Qmax * 0.72) - 0.4,
      label = "MR"
    ) +
    annotate(
      "text",
      x = 0.2,
      y = c0 + 0.35,
      label = expression(c[0]),
      hjust = 0
    ) +
    annotate(
      "text",
      x = 0.2,
      y = c1 + 0.35,
      label = expression(c[1]),
      hjust = 0
    ) +
    labs(title = title, x = "Q", y = "p") +
    coord_cartesian(xlim = c(0, Qmax), ylim = c(0, A)) +
    theme_minimal(base_size = 12)
}

pL <- make_panel(c0_L, c1_L, "Drastic process innovation (illustration)")
pR <- make_panel(c0_R, c1_R, "Non-drastic process innovation (illustration)")

# Print sequentially (Reveal will show them one after another on the slide)
pL
pR
``` -->

## Example: social planner (welfare comparison)

For linear demand $P(Q)=A-Q$ and marginal cost $c$:

Efficient output solves $P(Q)=c$:
$$
A - Q = c \quad \Rightarrow \quad Q^{sp}(c) = A - c.
$$

Total surplus (welfare) is the area under demand above cost:
$$
W(c) = \int_0^{A-c} (A - Q - c), dQ = \frac{(A-c)^2}{2}.
$$

Max WTP for the innovation (planner):
$$
\Delta W = W(c_1) - W(c_0) = \int_{c_1}^{c_0} (A-c), dc.
$$

## Replacement effect (intuition)

Key lecture point:

* A **monopolist** has positive pre-innovation profits, so innovation partly **replaces** existing profits.
* A **competitive firm** earns (near) zero pre-innovation profits, so innovation **creates** new profits.

Therefore, a competitive firm may be willing to pay more than a monopolist for the same innovation.

<!--
## Figure: Replacement effect (recreated)
 
```{r}
#| label: fig-replacement
#| fig-cap: "Replacement effect (recreation of lecture diagram; areas are illustrative)."

A <- 10
c0 <- 6
c1 <- 2

df <- tibble(Q = seq(0, A, length.out = 400)) |>
  mutate(P = A - Q, MR = A - 2 * Q)

# Monopoly outcomes before and after
Qm0 <- (A - c0) / 2
Pm0 <- (A + c0) / 2

Qm1 <- (A - c1) / 2
Pm1 <- (A + c1) / 2

# Demand-quantity at price=c0 (where competitive fringe would supply at cost)
Qc0 <- A - c0
Qc1 <- A - c1

# Construct illustrative rectangles/areas similar to the lecture shading
areas <- bind_rows(
  tibble(xmin = 0, xmax = Qm0, ymin = c0, ymax = Pm0, area = "1"),
  tibble(xmin = Qm0, xmax = Qc0, ymin = c0, ymax = Pm0, area = "2"),
  tibble(xmin = Qc0, xmax = Qm1, ymin = c1, ymax = Pm1, area = "3")
)

ggplot(df, aes(x = Q)) +
  geom_rect(
    data = areas,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    alpha = 0.25
  ) +
  geom_line(aes(y = P), linewidth = 1) +
  geom_line(aes(y = MR), linewidth = 0.9, linetype = "dashed") +
  geom_hline(yintercept = c0, linewidth = 0.8) +
  geom_hline(yintercept = c1, linewidth = 0.8) +
  geom_vline(xintercept = Qm0, linetype = "dotted") +
  geom_vline(xintercept = Qc0, linetype = "dotted") +
  geom_vline(xintercept = Qm1, linetype = "dotted") +
  geom_hline(yintercept = Pm0, linetype = "dotted") +
  geom_hline(yintercept = Pm1, linetype = "dotted") +
  annotate("text", x = Qm0 / 2, y = (c0 + Pm0) / 2, label = "1") +
  annotate("text", x = (Qm0 + Qc0) / 2, y = (c0 + Pm0) / 2, label = "2") +
  annotate("text", x = (Qc0 + Qm1) / 2, y = (c1 + Pm1) / 2, label = "3") +
  annotate(
    "text",
    x = 0.2,
    y = c0 + 0.35,
    label = expression(c[0]),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 0.2,
    y = c1 + 0.35,
    label = expression(c[1]),
    hjust = 0
  ) +
  annotate("text", x = A * 0.78, y = A - A * 0.78 + 0.4, label = "Demand") +
  annotate("text", x = A * 0.78, y = A - 2 * (A * 0.78) - 0.4, label = "MR") +
  labs(
    x = "Quantity",
    y = "Price",
    title = "Replacement effect (illustration)"
  ) +
  coord_cartesian(xlim = c(0, A), ylim = c(0, A)) +
  theme_minimal(base_size = 12)
```
 -->
## Exercise (Bertrand / process innovation)

Market:

* Inverse demand $P(Q) = 15 - Q$
* Bertrand competition
* Process innovation lowers MC from $c_0 = 3$ to $c_1 = 0$

Compare two market structures:

* Monopoly (before and after innovation)
* Perfect competition (before; monopoly rights/use after innovation)

Compute and compare:

* Maximum WTP for the innovation
* Change in total welfare (PS + CS) from innovation

## Incentives to innovate in oligopoly

Earlier ranking (under Bertrand framing in the lecture narrative):

* Max WTP: **Social planner > Perfect competition firm > Monopoly**

But in oligopoly:

* Incentives need not be monotone in “competition intensity”
* Key drivers:

  * number of firms
  * product differentiation
  * type of competition (price vs quantity)

## Impact of number of firms (Cournot intuition)

Linear Cournot with $n$ firms:

* Innovation incentives can follow an **inverse-U** as $n$ increases.

Two opposing effects as $n \uparrow$:

* **Competition effect**: profits for all firms fall
* **Competitive advantage effect**: a cost advantage becomes more valuable when there are more inefficient rivals

## Innovation affects market structure

Previously considered cases where:

* Monopoly remains monopoly even without innovation (no entry threat), or
* Innovation lets a non-monopolist become a monopolist.

Now: **incumbent monopoly threatened by entry**.

Innovation can change structure from monopoly to duopoly/oligopoly if the entrant acquires it.

## Monopoly threatened by entry: setup and payoffs

Setup:

* Incumbent and entrant can acquire a (minor) innovation reducing MC from $c_0$ to $c_1 < c_0$.
* Entrant can enter profitably **only if** it gets the innovation.

Payoffs:

* If incumbent gets innovation: entrant stays out

  * incumbent: $\pi^m(c_1)$
  * entrant: $0$

* If entrant gets innovation: entrant enters

  * incumbent: $\pi^d(c_0, c_1)$
  * entrant: $\pi^d(c_1, c_0)$

## Efficiency effect

Per-period value of innovation:

$$
V_I = \pi^m(c_1) - \pi^d(c_0, c_1), \qquad
V_E = \pi^d(c_1, c_0) - 0.
$$

Incumbent has higher incentives if:

$$
V_I > V_E
\iff
\pi^m(c_1) > \pi^d(c_1, c_0) + \pi^d(c_0, c_1),
$$

typically when products are **close substitutes**.

Interpretation (lecture):

* Fear of losing monopoly position strengthens incumbent’s incentive.
* Innovation may **prolong monopoly**.

## Seminal paper: Dasgupta & Stiglitz (1980)

Question: Does more firms imply more innovation?

Primitives:

* $U(Q)$ gross social benefit
* constant marginal cost $c$
* R&D expenditure $x$
* cost depends on R&D: $c=c(x)$ with $c'(x) < 0$

Social optimum:
$$
\max_{x,Q} V(x,Q) = U(Q) - c(x)Q - x
$$

FOC in $x$:
$$
-c'(x)Q = 1.
$$

## Free-entry oligopoly equilibrium conditions (DS 1980)

Firm problem (symmetric firms):
$$
\max_{x_i,q_i} \ \pi(x_i,q_i) = [P(q_i + q_{-i}) - c(x_i)]q_i - x_i
$$

FOC in $q_i$:
$$
\frac{\partial P}{\partial Q}q_i + P = c(x_i).
$$

Symmetric equilibrium: $q_1^*=\dots=q_N^* = Q^*/N$ and $x_1^*=\dots=x_N^* = x^*$, with
$$
P(Q^*)\left(1-\frac{\varepsilon}{N}\right) = c(x^*),
\qquad
\frac{1}{\varepsilon} = -\frac{\partial Q}{\partial P}\frac{P}{Q}.
$$

Free entry: $\pi=0$, implying
$$
(P-c)\frac{Q^*}{N^*} - x^* = 0
\quad\Rightarrow\quad
(P-c) = N^*\frac{x^*}{Q^*}.
$$

## Key implication: concentration and research intensity

Combining equilibrium conditions yields (lecture statement):

* No causal link: both are jointly determined.

Define “innovative potential”
$$
\alpha = -\frac{dc(x)}{dx}\frac{x}{c}.
$$

Then equilibrium concentration satisfies:
$$
\frac{1}{N^*} = \frac{1}{\varepsilon}\cdot \frac{\alpha}{1+\alpha}.
$$

## Dasgupta (1986): additional results (high level)

With $P(Q)=\sigma Q^{-\varepsilon}$ and $c(x)=\beta x^{-\alpha}$:

* Dynamic vs static efficiency trade-off:

  * $x^*(N+1) < x^*(N)$ but $Q^*(N+1) > Q^*(N)$
* Research intensity often maximized at moderate concentration (Scherer-style inverse-U)
* Greater innovative opportunities can be associated with higher concentration:
  $$
  \frac{\partial (1/N^*)}{\partial \alpha} > 0
  $$
* Demand growth ($\sigma$) stimulates R&D:
  $$
  \frac{\partial x^*}{\partial \sigma} > 0
  $$

## References & next week

References (as in the lecture):

* Belleflamme & Peitz (2015), *Industrial Organization: Markets and Strategies*, Ch. 18.1–18.2
* Dasgupta (1986), “The Theory of Technological Competition”
* Dasgupta & Stiglitz (1980), *Economic Journal* 90(358), 266–293
* Scotchmer (2004), *Innovation and Incentives*, Ch. 2–4

Next week: **Economics of patents and IPRs**
