---
title: "Exercise Set 1: Market Power and Welfare"
format: html
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

This exercise set covers two core models from **Lecture 1**:

- Cournot competition and welfare
- Hotelling linear city with a monopolist

These exercises are intended for **practice**. You can retry them and refresh the page to work through the logic again.

---

## Problem 1: Cournot duopoly and welfare

```{r}
#| results: asis
library(exams2forms)

cat(
  exams2forms(
    "exercises/exercises-rmd/cournot_duopoly_welfare.Rmd",
    n = 1
  ),
  sep = "\n"
)
```

---

## Problem 2: Hotelling linear city with monopoly

```{r}
#| results: asis
library(exams2forms)

cat(
  exams2forms(
    "exercises/exercises-rmd/hotelling_monopoly.Rmd",
    n = 1
  ),
  sep = "\n"
)
```

---

### How to use these exercises

* Enter your answers directly into the fields.
* Click **“Check”** to see whether your answers are correct.
* Use the feedback to understand where mistakes come from.
* These pages are **not graded** and are meant to support exam preparation.

If you want additional practice, the same models will reappear later in more general settings.
